dyngen: benchmarking with in silico single cells

### 2019-12-02 manuscript.Rmd
update

In the most simple case, $G$ is regulated by a single TF $H$, which can either be bound (state $S_0$) or unbound (state $S_1$). The probability $P(S_1)$ that $H$ is bound to the promoter region of $G$ is given by the abundance levels of $H$.

@@

>>>>

@@

\newcommand{\proptran}{f}
\newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
\newcommand{\yk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
\newcommand{\wi}[1]{\nu_#1}

The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$. Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.

### Propensity of transcription when $N=1$
In the simplest case when $N=1$, the promoter can be in one of two states. In state $S_0$, the promoter is not bound by any transcription factors, and in state $S_1$ the promoter is bound by $H_1$. Each state $S_j$ is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state. The propensity function is thus equal to the expected value of the activity of the promoter multiplied by the pre-mRNA production rate of $G$.

\begin{align}
  f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
\end{align}

For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient.

\begin{align}
  P(S_1) & = \frac{y_1^{n_1}}{k_1^{n_1} + y_1^{n_1}} \\
     & = \frac{(y_1/k_1)^{n_1}}{1 + (y_1/k_1)^{n_1}}
\end{align}

The Hill equation can be simplified by letting $\nu_i = \left(\frac{y_i}{k_i}\right)^{n_i}$.

\begin{align}
P(S_1) & = \frac{\nu_1}{1 + \nu_1} \label{eqn:hillsimp}
\end{align}

Since $P(S_0) = 1 - P(S_1)$, the activation function is formulated and simplified as follows.

\begin{align}
f(y_1) & = \text{wpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
   & = \text{wpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
   & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
\end{align}

### Propensity of transcription when $N=2$

When $N=2$, there are four states $S_j$. The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative). In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for two transcription factors.

Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:

\begin{align}
w_0 & = 1 \\
w_1 & = \nu_1 \\
w_2 & = \nu_2 \\
w_3 & = \nu_1 \cdot \nu_2
\end{align}

The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:

\begin{align}
    P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
       & = \frac{w_j}{1 + \nu_1 + \nu_2 + \nu_1 \cdot \nu_2} \\
       & = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting $P(S_j)$ and $w_j$ into $f(.)$ results in the following equation:

\begin{align}
f(y_1, y_2) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
 & = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
 & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
\end{align}

### Propensity of transcription for an arbitrary $N$
For an arbitrary $N$, there are $2^N$ states $S_j$. The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative). In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for $N$ transcription factors.

Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:

\begin{align}
  w_j & = \prod_{i=1}^{i \leq N} (j \text{ mod } i) = 1 \text{ ? } \nu_i \text{ : } 1
\end{align}

The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:

\begin{align}
P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
& = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting $P(S_j)$ into $f(.)$ yields:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
& = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
\end{align}

### Propensity of transcription for a large $N$
For large values of $N$, computing $f(.)$ is practically infeasible as it requires performing $2^N$ summations. In order to greatly simplify $f(.)$, $\alpha_j$ could be defined as 0 when one of the regulators inhibits transcription and 1 otherwise.

\begin{equation}
\alpha_j = \begin{cases}
 0 & \text{ if } \exists i : j \text{ mod } i = 1 \text{ and } H_i \text{ represses } G \\
 1 & \text{otherwise}
\end{cases} \label{eqn:assalpha}
\end{equation}

Substituting equation \ref{eqn:assalpha} into equation \ref{eqn:prop2n} and defining $R = \{1, 2, \ldots, N\}$ and $R^+ = \{i | H_i \text{ activates } G\}$ yields the simplified propensity function:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
\end{align}

### Independence, synergism and basal expression
The definition of $\alpha_j$ as in equation \ref{eqn:assalpha} presents two main limitations. Firstly, since $\alpha_0 = 1$, it is impossible to tweak the propensity of transcription when no transcription factors are bound. Secondly, it is not possible to tweak the independence and synergism of multiple regulators.

Let $\text{ba} \in [0,1]$ denote the basal expression strength $G$ (i.e. how much will $G$ be expressed when no transcription factors are bound), and $\text{sy} \in [0,1]$ denote the synergism of regulators $H_i$ of $G$, the transcription propensity becomes:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
\end{align}
